Analysis of $\piup$-$\piup$ lattice QCD scattering data

### 2015-03-16 Paper.tex
Write a bit more

\documentclass[11pt, english, fleqn, DIV=15, headinclude]{scrartcl}

@@

\begin{abstract}
    % TODO
\end{abstract}

@@

\subsection{Metropolis algorithm}

\subsection{Correlation functions}

@@

In order to get rid of all those problems at the same time one does a meta
analysis of the data itself. To get back to the example of the mean, let $X$ be
the set of $N$ data points ($x_i \in \R$). Then the estimator of the mean
$\bar\mu$ can be computed directly from $X$. Now $R$ bootstrap samples $X^r$
are computed from the set by randomly selecting $N$ elements from $X$ and
putting them into $X^r$ (upper index, no power). Since $X$ is sampled from the
original distribution, it resembles the original distribution as good as it can
with only $N$ elements. $X^r$ therefore is sampled from an approximation of the
original distribution. Then the estimators for the means $\bar\mu^r$ are
computed from the bootstrapped sets. The mean of all the $\bar\mu^r$ should
give $\bar\mu$, which was computed from the original data. The error estimate
is then the standard deviation of all the $\bar\mu^r$.

@@

The pion masses are contained in the correlation functions in the form of an
exponential decay constant. Since time is periodic in this simulation, the
correlation function is also periodic with the temporal lattice extent $T$.
Therefore, the expectation is not a simple exponential decay—$\exp(-\lambda
t)$—but a cosh-like function. The four-point functions also have a constant
contribution due to the finite lattice extent. For both two and four point
functions, the functions that I fitted to the folded data with parameters $\vec
\lambda$ are:
\begin{equation}
    \label{eq:fit2}
    m_\piup(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])}
\end{equation}
and
\begin{equation}
    \label{eq:fit4}
    m_{\piup\piup}(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])} + \lambda_3.
\end{equation}

@@

\begin{figure}[htbp]
    \centering
    \begin{minipage}[c]{0.49\linewidth}
        \centering
        \includegraphics[width=\linewidth]{plots/A100_24_L24_T48_beta190_mul0100_musig150_mudel190_kappa1632550__ev120__TB2_SO_LI6_new_c2_folded.pdf}
    \end{minipage}
    \hfill
    \begin{minipage}[c]{0.49\linewidth}
        \centering
        \includegraphics[width=\linewidth]{plots/A100_24_L24_T48_beta190_mul0100_musig150_mudel190_kappa1632550__ev120__TB2_SO_LI6_new_c4_folded.pdf}
    \end{minipage}
    \caption{%
        Folded correlation functions with cosh-like fit function using a
        logarithmic ordinate scale. On the left side, the two point function is
        shown, the four point function on the right. Only the blue data points
        were used for fitting. The right ordinate shows the residuals in great
        magnification (red). One can see that the errors seem way too large.
        This is a result of the autocorrelation of the data points.
    }
    \label{fig:folded}
\end{figure}

The correlation functions themselves are highly correlated with respect to
time. Therefore, a regular least squared fit would give a $\chi^2$ that would
be way too low. The $p$-values usually end up around 1, which does not imply a
perfect fit but rather that the data fits the model too \emph{well}. This, in
turn, means that errors and residuals are over-estimated with the given model.
A different model is called for: the correlated fit.

For the correlated fit, a new likelihood function is needed. I chose to keep
the least squared likelihood function but incorporate the correlation into the
$\chi^2$. First, a correlation matrix $\tens C$ is needed, which is computed
from the $R$ bootstrap samples:
\begin{equation}
    \label{eq:correlation_matrix}
    C_{ij} := \frac{1}{R[R-1]} \sum_{k=1}^R
    [x_{ik} - \bar x_{iR}] [x_{jk} - \bar x_{jR}]
    \eqnsep
    \bar x_{iR} := \frac 1R \sum_{r=1}^R x_{ir}.
\end{equation}

@@

In the curve fitting process, a function like
\texttt{scipy.optimize.curve\_fit} tries to find the parameters $\lambda$ such
that $\chi^2$ becomes minimal. In my implementation, I used
\texttt{scipy.optimize.leastsq} which tries to minimize the squared norm of a
vector valued function. With a Cholesky decomposition, this can be done like
so: Define the square bracket in Eq.~\eqref{eq:chisq} to be the residual vector
$\vec r$. Then the matrix multiplication can be written as
\begin{equation}
    \chi^2 = \vec r^\text T \tens C^{-1} \vec r.
\end{equation}
The Cholesky decomposition gives me $\tens C^{-1} = \tens U^\dagger \tens U$,
where $\tens U$ is a upper triangle matrix. Then I can write
\begin{equation}
    \chi^2 = [\tens U \vec r]^\dagger [\tens U \vec r]
    = \| \tens U \vec r \|^2
\end{equation}
where $\tens U \vec r$ is a vector that can be fed into
\texttt{scipy.optimize.leastsq}.

@@

Using the computed $\mpi$ and $\mpipi$ with Lüscher's formula, I can
numerically solve for the scattering length $a_0$. In my program, I use
\texttt{scipy.optimize.brentq} that is based on the algorithm by Brent (1973).

The computed scattering lengths are shown in Table~\ref{tab:masses}.

@@

\begin{table}
    \centering
    \begin{tabular}{lSSS}
        ensemble & {$m_{\piup}$}  & {$m_{\piup\piup}$} & {$a_0$} \\
        \midrule
        A100.24 & 0.22238 +- 0.00023 & 0.45125 +- 0.00052 & -1.346 +- 0.029 \\
        A100.24 & 0.22233 +- 0.00040 & 0.45083 +- 0.00095 & -1.287 +- 0.132 \\
        A100.24 & 0.22239 +- 0.00024 & 0.45111 +- 0.00053 & -1.316 +- 0.039 \\
        A30.32  & 0.12416 +- 0.00055 & 0.25135 +- 0.00144 & -0.904 +- 0.273 \\
        A40.20  & 0.14779 +- 0.00078 & 0.31429 +- 0.00162 & -1.426 +- 0.052 \\
        A40.24  & 0.14447 +- 0.00052 & 0.29815 +- 0.00109 & -1.255 +- 0.051 \\
        A40.24  & 0.14453 +- 0.00031 & 0.29842 +- 0.00071 & -1.272 +- 0.051 \\
        A40.32  & 0.14126 +- 0.00022 & 0.28626 +- 0.00055 & -1.228 +- 0.092 \\
        A60.24  & 0.17275 +- 0.00052 & 0.35278 +- 0.00126 & -1.194 +- 0.112 \\
        A60.24  & 0.17279 +- 0.00048 & 0.35404 +- 0.00099 & -1.361 +- 0.091 \\
        A80.24  & 0.19930 +- 0.00024 & 0.40511 +- 0.00057 & -1.228 +- 0.043 \\
        B55.32  & 0.15553 +- 0.00022 & 0.31589 +- 0.00055 & -1.676 +- 0.117 \\
        D45.32  & 0.12047 +- 0.00046 & 0.25057 +- 0.00137 & -2.416 +- 0.205
    \end{tabular}
    \caption{%
        Computed masses from correlation functions. The last column shows the
        scattering length $a_0$ which is computed using Lüscher's formula,
        Eq.~\eqref{eq:luescher}.
    }
    \label{tab:masses}
\end{table}

>>>>

\documentclass[11pt, english, fleqn, DIV=10, headinclude]{scrartcl}

@@

\begin{abstract}
    Correlation functions of pions that were computed with lattice QCD methods
    are analyzed. The effective masses of single pions and two interacting
    pions are extracted. Using a finite size formula by Lüscher, the S-wave
    scattering length is computed.
\end{abstract}

@@

% Weighted generation with HMC

% Downloadable ensembles

% Operators on data

% Contractions

@@

In order to get rid of all those problems at the same time one does a meta
analysis of the data itself. To get back to the example of the mean, let $X$ be
the set of $N$ data points ($x_i \in \R$). Then the estimator of the mean
$\bar\mu$ can be computed directly from $X$. Now $R$ bootstrap samples $X^r$
(upper index, no power) are computed from the set by randomly selecting $N$
elements from $X$ and putting them into $X^r$. Since $X$ is sampled from the
original distribution, it resembles the original distribution as good as it can
with only $N$ elements. $X^r$ therefore is sampled from an approximation of the
original distribution. Then the estimators for the means $\bar\mu^r$ are
computed from the bootstrapped sets. The mean of all the $\bar\mu^r$ should
give $\bar\mu$, which was computed from the original data. The error estimate
is then the standard deviation of all the $\bar\mu^r$.

@@

The pion masses are contained in the correlation functions in the form of an
exponential decay constant. Since time is periodic in this simulation, the
correlation function is also periodic with the temporal lattice extent $T$.
Therefore, the expectation is not a simple exponential decay, $\exp(-\lambda
t)$, but a cosh-like function. The four-point functions also have a constant
contribution due to the finite lattice extent. For both two and four point
functions, the functions that I fitted to the folded data with parameters $\vec
\lambda$ are:
\begin{equation}
    \label{eq:fit2}
    m_\piup(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])}
\end{equation}
and
\begin{equation}
    \label{eq:fit4}
    m_{\piup\piup}(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])} + \lambda_3.
\end{equation}

@@

\begin{figure}[htbp]
    \centering
    \begin{minipage}[c]{0.49\linewidth}
        \centering
        \includegraphics[width=\linewidth]{plots/A100_24_L24_T48_beta190_mul0100_musig150_mudel190_kappa1632550__ev120__TB2_SO_LI6_new_c2_folded.pdf}
    \end{minipage}
    \hfill
    \begin{minipage}[c]{0.49\linewidth}
        \centering
        \includegraphics[width=\linewidth]{plots/A100_24_L24_T48_beta190_mul0100_musig150_mudel190_kappa1632550__ev120__TB2_SO_LI6_new_c4_folded.pdf}
    \end{minipage}
    \caption{%
        Folded correlation functions with cosh-like fit function. On the left
        side, the two point function is shown using a logarithmic ordinate
        scale. The four point function is shown on the right. Only the blue
        data points were used for fitting. The right ordinate shows the
        residuals in great magnification (red). One can see that the errors
        seem way too large compared to the residuals. This is a result of the
        autocorrelation of the data points.
    }
    \label{fig:folded}
\end{figure}

The correlation functions themselves are highly correlated with respect to
time. Therefore, a regular least squared fit would give a $\chi^2$ that would
be way too low for the assumed degrees of freedom. The $p$-values usually end
up around 1, which does not imply a perfect fit but rather that the data fits
the model \emph{too well}. This, in turn, means that errors and residuals are
over-estimated with the given model. A different model is called for: the
correlated fit.

For the correlated fit, a new likelihood function is needed. I chose to keep
the least squared likelihood function but incorporate the correlation into the
$\chi^2$. First, a correlation matrix $\tens C$ is needed, which is computed
from the $R$ bootstrap samples:
\begin{equation}
    \label{eq:correlation_matrix}
    C_{ij} := \frac{1}{R[R-1]} \sum_{r=1}^R
    [x_{ir} - \bar x_{iR}] [x_{jr} - \bar x_{jR}]
    \eqnsep
    \bar x_{iR} := \frac 1R \sum_{r=1}^R x_{ir}.
\end{equation}

@@

In the curve fitting process, a function like
\texttt{scipy.optimize.curve\_fit} tries to find the parameters $\lambda$ such
that $\chi^2$ becomes minimal. In my implementation, I used
\texttt{scipy.optimize.leastsq} which tries to minimize the squared norm of a
vector valued function. With a Cholesky decomposition, this can be done like
so: Define the square bracket in Equation~\eqref{eq:chisq} to be the residual
vector $\vec r$. Then the matrix multiplication can be written as
\begin{equation}
    \chi^2 = \vec r^\text T \tens C^{-1} \vec r.
\end{equation}
The Cholesky decomposition gives me $\tens C^{-1} = \tens U^\dagger \tens U$,
where $\tens U$ is a upper triangle matrix. Then I can write
\begin{equation}
    \chi^2 = [\tens U \vec r]^\dagger [\tens U \vec r]
    = \| \tens U \vec r \|^2
\end{equation}
where $\tens U \vec r$ is a vector that can be fed into
\texttt{scipy.optimize.leastsq}.

@@

Using the computed $\mpi$ and $\mpipi$ with Lüscher's formula, I can
numerically solve for the scattering length $a_0$. In my program, I use
\texttt{scipy.optimize.brentq} that is based on the algorithm by Brent (1973).
The computed scattering lengths are shown in Table~\ref{tab:masses}.

Equation~\eqref{eq:luescher} can be motivated like so: Let $\hat H = \hat H_0 +
\hat V$ be the Hamiltonian for two identical particles where $\hat H_0$ is the
free Hamiltonian and $\hat V$ a spherically symmetric potential. The
eigenfunctions of $\hat H_0$ are plane waves which are denoted $\ket{\vec p,
\vec q}$. In continous space, the normalization is
\parencite[(2.10)]{luescher/volume_dependence}
\[
    \braket{\vec p', \vec q'|\vec p, \vec q}
    = [2\piup]^6 \sbr{
        \delta(\vec p' - \vec p)
        \delta(\vec q' - \vec q)
        +
        \delta(\vec p' - \vec q)
        \delta(\vec q' - \vec p)
    }.
\]
On the discrete lattice, the possible momentum values are restricted to lattice
values and the potential will be periodic due to periodic boundary conditions.
This changes the normalization to
\parencite[(2.21)]{luescher/volume_dependence}
\[
    \braket{\vec p', \vec q'|\vec p, \vec q}
    = L^6 \sbr{
        \delta_{\vec p', \vec p}
        \delta_{\vec q', \vec q}
        +
        \delta_{\vec p', \vec q}
        \delta_{\vec q', \vec p}
    }.
\]

The matrix element of the potential $\vec V$ are given by
\parencite[(2.21)]{luescher/volume_dependence}
\[
    \braket{\vec p', \vec q'| \hat V | \vec p, \vec q}
    = L^3 \delta_{\vec P', \vec P} \hat V\del{\frac12 [\vec p' - \vec q'],
    \frac12 [\vec p - \vec q]},
\]
where $\vec P$ is the total momentum of the two particles. The energy shift in
first order of $V$ for the ground state (all momenta are zero) then is the
quotient \parencite[(2.24)]{luescher/volume_dependence}:
\[
    \Deltaup E = \frac{1}{2L^3} \hat V(\vec 0, \vec 0) + \mathcal O(V^2).
\]
The scattering amplitude as well as the S-wave scattering length $a_0$ can be
expanded in powers of $\hat V$ (Born series). Using first order term from 
\parencite[(2.18)]{luescher/volume_dependence}, above equation can be rewritten
with the scattering length
\parencite[(2.25)]{luescher/volume_dependence}:
\[
    \Deltaup E = - \frac{4\piup a_0}{m L^3} + \mathcal O(V^2).
\]

To first order on $V$, the leading term scales with $L^{-3}$, like in
Equation~\eqref{eq:luescher}. To obtain more terms, a complete perturbation
expansion needs to be written out. That expansion will be both in terms of
$L^{-1}$ \emph{and} $V$. \citeauthor{luescher/volume_dependence} then was able
to group those terms in powers of $L^{-1}$ and obtain
Equation~\eqref{eq:luescher}.

% FIXME Maybe write a bit more on the methods?

@@

My analysis of the correlation functions gives me values for the masses of a
single pion ($\mpi$) and a pair of them ($\mpipi$). Using
Equation~\eqref{eq:luescher}, I was able to compute the S-wave scattering
length $a_0$ from the energy difference. Those values are listed in
Table~\ref{tab:masses}.

\begin{table}
    \centering
    \begin{tabular}{lSSS}
        ensemble & {$m_{\piup}$}  & {$m_{\piup\piup}$} & {$a_0$} \\
        \midrule
        A100.24 & 0.22238 +- 0.00023 & 0.45125 +- 0.00052 & -1.346 +- 0.029 \\
        A100.24 & 0.22233 +- 0.00040 & 0.45083 +- 0.00095 & -1.287 +- 0.132 \\
        A100.24 & 0.22239 +- 0.00024 & 0.45111 +- 0.00053 & -1.316 +- 0.039 \\
        A30.32  & 0.12416 +- 0.00055 & 0.25135 +- 0.00144 & -0.904 +- 0.273 \\
        A40.20  & 0.14779 +- 0.00078 & 0.31429 +- 0.00162 & -1.426 +- 0.052 \\
        A40.24  & 0.14447 +- 0.00052 & 0.29815 +- 0.00109 & -1.255 +- 0.051 \\
        A40.24  & 0.14453 +- 0.00031 & 0.29842 +- 0.00071 & -1.272 +- 0.051 \\
        A40.32  & 0.14126 +- 0.00022 & 0.28626 +- 0.00055 & -1.228 +- 0.092 \\
        A60.24  & 0.17275 +- 0.00052 & 0.35278 +- 0.00126 & -1.194 +- 0.112 \\
        A60.24  & 0.17279 +- 0.00048 & 0.35404 +- 0.00099 & -1.361 +- 0.091 \\
        A80.24  & 0.19930 +- 0.00024 & 0.40511 +- 0.00057 & -1.228 +- 0.043 \\
        B55.32  & 0.15553 +- 0.00022 & 0.31589 +- 0.00055 & -1.676 +- 0.117 \\
        D45.32  & 0.12047 +- 0.00046 & 0.25057 +- 0.00137 & -2.416 +- 0.205
    \end{tabular}
    \caption{%
        Computed masses from correlation functions. The last column shows the
        scattering length $a_0$ which is computed using Lüscher's formula,
        Equation~\eqref{eq:luescher}.
    }
    \label{tab:masses}
\end{table}

The different ensembles assume different masses for the quarks. The first
number behind the letter gives the assumed quark mass. In most cases, a lower
quark mass also yields a lower pion mass $\mpi$. Those pion masses are not
directly the physical ones, one needs to extrapolate to the physical point. The
masses and scattering lengths are given in lattice units and depends on the
lattice spacing. The product of mass and scattering length, $a_0 \mpi$, is
independent of the lattice spacing and a good measure for extrapolation.
Table~\ref{tab:computed} shows this product, as well as the decay constant.
